Taking uncertainty in word embedding bias estimation seriously: a  bayesian

### 2023-05-25 paper/paperDraft.Rmd
revised visualizations, fixed intro

Natural language processing (NLP) models are used to perform various language-related tasks such as providing email filters, intelligent assistants, search results,  translations, text analysis and so on. Such tools require input words to be represented by  vectors of real numbers. This can usually be done with word embeddings.  A common way to construct an embedding is to use  a large natural language corpus to train  a  neural network to assign vectors to words that optimize for co-occurrence prediction accuracy. The vectors can then be compared in terms of their similarity -- the usual measure is cosine similarity -- and the results of such comparisons can be used in downstream tasks. Roughly speaking, cosine similarity is an imperfect mathematical proxy for semantic similarity [@Mikolov2013efficient].

It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that  such models can learn implicit biases that reflect harmful stereotypical thinking---for example,  the (vector corresponding to the) word \textit{she} might be  much closer in the vector space to  the word \textit{cooking} than  the word \textit{he}. Such phenomena may be undesirable at least  in some  downstream tasks, such as web search, recommendations, and so on. To investigate such issues, several measures of bias in word embeddings have been formulated and applied. Our goal is to use two prominent examples of such measures to argue that this approach oversimplifies the situation and to develop a Bayesian alternative.

@@

Another response might be to insist that there is no bias introduced by the use of machine learning methods here, since the algorithm is simply learning to correctly predict co-occurrences based on what "reality" looks like. However, this objection overlooks the fact that we, humans, are the ones who construct this linguistic reality, is shaped in part by the natural language processing tools we use on a massive scale. Sure, if there is unfairness and our goal is to diagnose it, we should do complete justice to learning it in the model used in a study thereof. One example of this approach is [@Garg2017hundredYears]. However, if our goal is to develop downstream tools that perform tasks that we care about without further perpetuating  or exacerbating harmful stereotypes, we still have good reasons to try to minimize the negative impact, at least as long as this does not damage the overall performance of the tool. Moreover, it is often not the case that the corpora mirror reality---to give a trivial example, heads are spoken of more often than kidneys, but this does not mean that kidneys occur much less often in reality than heads. To give a more relevant example, the disproportionate association of female words with female occupations in a corpus actually greatly exaggerates the actual lower disproportion in the real distribution of occupations. [@gordon2012reporting].

In what follows, we focus on  two popular  measures of bias applicable to many existing word embeddings ( \emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.} \emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.} and \emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}): \emph{Word Embedding Association Test}  (\textsf{WEAT}) [@Caliskan2017semanticsBiases],  and \emph{Mean Average Cosine Distance} (\textsf{MAC})  [@Manzini2019blackToCriminal].  We first explain how these measures are supposed to work. Then we argue that they are problematic for various reasons, a key one being that by pre-averaging data they manufacture false confidence, which we illustrate in terms of simulations showing that the measures often suggest the existence of bias even if by design it is non-existent in a simulated dataset.

We  propose to replace them with a Bayesian data analysis (hierarchical models), which not only provides more modest and realistic assessment of the uncertainty involved, but also allows for inspection at various levels of granularity. Once we introduce the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC} (and the problem generalizes to any approach that focuses on chasing a single numeric metric): the word list sizes and sample sizes used in the studies are usually small,^[Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25.]  posterior density intervals are fairly wide,   often the differences between associated, different and neutral human predicates, are not very impressive. Also, a preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable. The tools that we have proposed, however, allow for a more fine-grained and multi-level evaluation of bias and debiasing in language models without abandoning modesty about the uncertainties involved.

# Two measures of bias: WEAT and MAC

Conceptually the idea is that if a particular harmful stereotype is learned in a particular embedding, then certain groups of words will be systematically closer to (or further from) each other. This gives rise to the idea of protected groups---for example, in guiding online search completion or recommendation, female words might require protection in that they should not be systematically closer to stereotypically female job names, and male words require protection in that they should not be systematically closer to toxic masculinity stereotypes, such as "tough", "never complaining" or "macho".^[However, for some research-related purposes, such as the study of stereotypes across history [@Garg2017hundredYears], embeddings which do not protect certain classes may also be useful.]

The key role in the measures to be discussed is played by the notion of cosine distance (or, symmetrically, by cosine similarity). These are defined as follows:^[Here, "$-$" stands for point-wise difference, "$\cdot$" stands for the dot product operation, and $\lVert a\rVert  = \sqrt{(a \cdot a)}$.]
\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
\\
\tag{Distance}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B).
\end{align}
Note  that this terminology is slightly misleading, as mathematically cosine distance is not a distance measure, because it does not satisfy the triangle inequality, as generally $\mathsf{cosineDistance}(A,C) \not \leq \mathsf{cosineDistance}(A,B)+ \mathsf{cosineDistance}(B,C)$. We will keep using this mainstream terminology.

@@

\begin{tabular}[t]{lllr}
\toprule
protected words ($T$) & attributes & attribute set ($A_j$) & cosine distance\\
\midrule
\cellcolor{gray!15}{rabbi} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{1.0306175}\\
church & familial & christianStereotype & 0.7087424\\
\cellcolor{gray!15}{synagogue} & \cellcolor{gray!15}{liberal} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{0.7922607}\\
jew & familial & christianStereotype & 0.9783245\\
\cellcolor{gray!15}{quran} & \cellcolor{gray!15}{dirty} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{1.1207093}\\
muslim & uneducated & muslimStereotype & 0.5160429\\
\cellcolor{gray!15}{torah} & \cellcolor{gray!15}{terrorist} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{0.9341137}\\
quran & hairy & jewStereotype & 1.1764642\\
\cellcolor{gray!15}{synagogue} & \cellcolor{gray!15}{violent} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{0.9549743}\\
bible & cheap & jewStereotype & 1.2234364\\
\cellcolor{gray!15}{christianity} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{0.9728545}\\
muslim & hairy & jewStereotype & 0.8788219\\
\cellcolor{gray!15}{islam} & \cellcolor{gray!15}{critical} & \cellcolor{gray!15}{christianStereotype} & \cellcolor{gray!15}{0.7880706}\\
muslim & conservative & christianStereotype & 0.4453191\\
\cellcolor{gray!15}{mosque} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{1.1541524}\\
\bottomrule
\end{tabular}

@@

# Methodological problems with cosine-based bias metrics

@@

In Figure  \ref{fig:empirical} we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine similarity rather than cosine distance; this is a trivial linear transformation) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, \textsf{MAC} was .859 and considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of 1-\textsf{MAC} deserving debiasing according to the authors is quite high.

```{r cosineDistributions6b,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "75%", warning = FALSE, message = FALSE}

@@

violinReligionReddit <- ggplot(religionReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+
  geom_vline(xintercept = .14) + geom_hline(yintercept = -.14, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 10)+ geom_hline(yintercept = .14, color = "grey")+
  labs(title =  "Religion (Reddit)", subtitle = "Empirical distribution of cosine similarities")+
  geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "56%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "55%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "43%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "36%")+
  theme(plot.title.position = "plot",legend.position="top")

violinGenderReddit <- ggplot(genderReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  geom_vline(xintercept = .38) + geom_hline(yintercept = -.38, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 10)+ geom_hline(yintercept = .38, color = "grey")+
  ggtitle("Gender: empirical distribution of cosine similarity")+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "96%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "87%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "53%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "47%")+
  labs(title =  "Gender (Reddit)", subtitle = "Empirical distribution of cosine similarities")+
  theme(plot.title.position = "plot",legend.position='none')

violinRaceReddit <- ggplot(raceReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  geom_vline(xintercept = .11) + geom_hline(yintercept = -.11, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 10)+ geom_hline(yintercept = .11, color = "grey")+
   labs(title =  "Race (Reddit)", subtitle = "Empirical distribution of cosine similarities")+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "46%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "67%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "39%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "33%")+
  theme(plot.title.position = "plot",legend.position='none')

```
\pagebreak

\begin{figure}[H]
```{r fig:cosineDistributions8b,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", fig.height = 16, warning = FALSE, message = FALSE}

grid.arrange(violinReligionReddit, violinGenderReddit, violinRaceReddit, ncol = 1)

@@

\caption{Empirical distributions of cosine similarities (1-distances) for three word lists used in  the original paper.  }

\label{fig:empirical}
\end{figure}

@@

#  Pre-averaging, bootstrapping and testing with the null model

The approaches we have been describing use  means of mean average cosine similarities to measure  similarity between protected words and attributes coming from harmful stereotypes. However, a closer inspection look at the individual values that are taken for the calculations, it turns out that there are quite a few  outliers and surprisingly dissimilar words. This problem becomes transparent when we examine the visualizations of the individual cosine distances, following the idea that one of the first steps in understanding data is to look at it.

@@

visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 11)
```

```{r visPriestd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

@@

\caption{Actual distances for two protected words. For the protected word "muslim", the most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. For the word "priest", the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.}
\end{figure}

@@

ourESplot <- ggplot(dataPLOT )+geom_histogram(aes(x=biasesD0, y = ..density..), bins = 50, alpha = .6)+
  theme_tufte(base_size = 10)+
  labs(title="Distribution of effect sizes given a null model",
       subtitle = "Simulation with Norm(0,0.8), n=10k")+ xlab("effect size")+
    theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  geom_vline(xintercept = ourEffectSize, linewidth = .5,
             lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  annotate("label", x = ourEffectSize, y = .06,
           label =  paste(round(ourEffectSize,2),
                          ", p=", 0.38, sep = ""), size = 3)

ourStatPlot <- ggplot(dataPLOT )+ geom_histogram(aes(x=svaluesD0, y = ..density..), alpha = .6, bins = 50) +
  theme_tufte(base_size = 10)+
  labs(title="Distribution of test statistics given a null model",
       subtitle = "Simulation with Norm(0,0.8), n=10k")+ xlab("test statistic")+
    theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank()) +
  geom_vline(xintercept = ourStatistic, linewidth = .5,
             lty = 2, alpha = .5)+theme_tufte(base_size = 10)+
  annotate("label", x = ourStatistic, y = 1.2,
           label =  paste(round(ourStatistic,2),
                          ", p=", 0.01, sep = ""), size = 3)

@@

\begin{figure}[H]
```{r fig:ourDistancesPlot2,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%", warning = FALSE, message = FALSE}

@@

\caption{Distributions of test statistics and effect sizes based on 10k simulations on the assumption of a null model in which all distances come from normal distribution with $\mu =0, \sigma = .08$.}
\label{fig:ourDistances}
\end{figure}

@@

# Bayesian method

## Introduction and model definition

@@

## Posterior predictive check

@@

\begin{figure}[H]
```{r figposteriorPrCheckA6b,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%", dpi = 300, warning = FALSE, message = FALSE}
grid.arrange(posteriorPlotsA[[3]], posteriorPlotsB[[3]], ncol = 2)
```
\caption{Example of a posterior predictive check. (Left) Actual cosine distances are plotted against mean predictions with 89\% highest posterior density intervals. Notice that 90\% of actual values fall within the 89\% HPDI and 55\% of actual values fall into 50\% HPDI, which indicates appropriate performance of the model. The left-right alignment of differrent colors coressponds to the fact that cosine differences between elements of different categories differ, to some extent systematically (this will be studied in the results section). (Right) Densities of predicted and observed distances.}
\label{fig:posteriorCheck1}
\end{figure}

@@

# Results and discussion

>>>>

It has been suggested^[See for instance [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal].]  that  language models can learn implicit biases that reflect harmful stereotypical thinking---for example,  the (vector corresponding to the) word \textit{she} might be  much closer in the vector space to  the word \textit{cooking} than  the word \textit{he}. Such phenomena are undesirable at least  in some  downstream tasks, such as web search, recommendations, and so on. To investigate such issues, several measures of bias in word embeddings have been formulated and applied. Our goal is to use two prominent examples of such measures to argue that this approach oversimplifies the situation and to develop a Bayesian alternative.

A common approach in natural language processing is to  represent words  by  vectors of real numbers---such representations are called \emph{embeddings}.  One way to construct an embedding---we will focus our attention on non-contextual language models^[One example of a contextualized representation is BERT. Another is GPT.]---is to use  a large corpus to train  a  neural network to assign vectors to words in a way that optimizes for co-occurrence prediction accuracy. Such vectors can then be compared in terms of their similarity---the usual measure is cosine similarity---and the results of such comparisons can be used in downstream tasks. Roughly speaking, cosine similarity is an imperfect mathematical proxy for semantic similarity [@Mikolov2013efficient].

@@

Another response might be to insist that there is no bias introduced by the use of machine learning methods here, since the algorithm is simply learning to correctly predict co-occurrences based on what "reality" looks like. However, this objection overlooks the fact that we, humans, are the ones who construct this linguistic reality, which is shaped in part by the natural language processing tools we use on a massive scale. Sure, if there is unfairness and our goal is to diagnose it, we should do complete justice to learning it in the model used to study it. One example of this approach is [@Garg2017hundredYears], where the authors use language  models to study the shape of certain biases across a century.

However, if our goal is to develop downstream tools that perform tasks that we care about without further perpetuating  or exacerbating harmful stereotypes, we still have good reasons to try to minimize the negative impact. Moreover, it is often not the case that the corpora mirror reality---to give a trivial example, heads are spoken of more often than kidneys, but this does not mean that kidneys occur much less often in reality than heads. To give a more relevant example, the disproportionate association of female words with female occupations in a corpus actually greatly exaggerates the actual lower disproportion in the real distribution of occupations [@gordon2012reporting].

In what follows, we focus on  two popular  measures of bias applicable to many existing word embeddings, such as \emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.} \emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.} and \emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}: \emph{Word Embedding Association Test}  (\textsf{WEAT}) [@Caliskan2017semanticsBiases],  and \emph{Mean Average Cosine Distance} (\textsf{MAC})  [@Manzini2019blackToCriminal].  We first explain how these measures are supposed to work. Then we argue that they are problematic for various reasons---the key one being that by pre-averaging data they manufacture false confidence, which we illustrate in terms of simulations showing that the measures often suggest the existence of bias even if by design it is non-existent in a simulated dataset.

We  propose to replace them with a Bayesian data analysis, which not only provides more modest and realistic assessment of the uncertainty involved, but in which hierarchical models allow for inspection at various levels of granularity. Once we introduce the method, we apply it to multiple word embeddings and results of supposed debiasing, putting forward some general observations that are not exactly in line with the usual picture painted in terms of \textsf{WEAT} or \textsf{MAC}.

Most of the  problems that we point out generalize to any existing approach that focuses on chasing a single numeric metric of bias. (1) They treat the results of pre-averaging as raw data in statistical significance tests, which in this context is bound to overestimate significance. We show similar results can easily be obtained when sampling from null models with no bias. (2) The word list sizes and sample sizes used in the studies are usually small,^[Depending on a list for [@Caliskan2017semanticsBiases] the range for protected words is between 13 and 100, and for attributes between 16 and 25; for [@Manzini2019blackToCriminal] the range for protected words is between 14 and 18, and for attributes between 11 and 25.] 
(3) Many studies do not use any control predicates, such as random neutral words or neutral human predicates for comparison.

On the constructive side, we develop and deploy our method, and the results are, roughly, as follows. (A) Posterior density intervals are fairly wide and    the average differences between associated, different and neutral human predicates, are not very large. (B) A preliminary inspection suggests that  the desirability of changes obtained by the usual debiasing methods is debatable.

In Section \ref{sec:two} we describe the two key measures discussed in this paper, \textsf{WEAT} and \textsf{MAC}, explaining how they are calculated and how they are supposed to work. In Section \ref{sec:challenges} we first argue in Subsection \ref{subsec:interpretability}, that it is far from clear how results given in terms of \textsf{WEAT} or \textsf{MAC} are to be interpreted. Second, in Subsection \ref{subsec:problems} we explain the statistical problems that arise when one uses pre-averaged data in such contexts, as these measures do. In Section \ref{sec:bayesian} we explain the alternative Bayesian approach that we propose. In Section \ref{sec:results} we elaborate on the results that it leads to, including a somewhat skeptical view of the efficiency of debiasing methods, discussed in Subsection \ref{subsec:rethinking}. Finally, in Section \ref{sec:related} we spend some time placing our results in the ongoing discussions.^[\textbf{Disclaimer:} throughout the paper we will be mentioning and using word lists and stereotypes we did not formulate, which does not mean we condone any judgment made therein or underlying a given word selection. For instance, the Gender dataset does not recognize non-binary categories, and yet we use it without claiming that such categories should be ignored.]

# Two measures of bias: WEAT and MAC
\label{sec:two}

Conceptually, the idea is that if a particular harmful stereotype is learned in a particular embedding, then certain groups of words will be systematically closer to (or further from) each other. This gives rise to the idea of protected groups---for example, in guiding online search completion or recommendation, female words might require protection in that they should not be systematically closer to stereotypically female job names, such as "nurse", "librarian", "waitress", and male words require protection in that they should not be systematically closer to toxic masculinity stereotypes, such as "tough", "never complaining" or "macho".^[However, for some research-related purposes, such as the study of stereotypes across history [@Garg2017hundredYears], embeddings which do not protect certain classes may also be useful.]

The key role in the measures to be discussed is played by the notion of cosine distance (or, symmetrically, by cosine similarity). These are defined as follows:^[Here, "$-$" stands for point-wise difference, "$\cdot$" stands for the dot product operation, and $\lVert a\rVert  = \sqrt{(a \cdot a)}$.] $^{\!\!\!  , \,}$^[Note that this terminology is slightly misleading, as mathematically cosine distance is not a distance measure, because it does not satisfy the triangle inequality, as generally $\mathsf{cosineDistance}(A,C) \not \leq \mathsf{cosineDistance}(A,B)+ \mathsf{cosineDistance}(B,C)$. We will keep using this mainstream terminology.]
\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
\\
\tag{Distance}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B).
\end{align}

@@

\begin{tabular}[t]{lllr}
\toprule
protected words ($T$) & attributes & attribute set ($A_j$) & cosine distance\\
\midrule
\cellcolor{gray!15}{rabbi} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{1.03}\\
church & familial & christianStereotype & 0.70\\
\cellcolor{gray!15}{synagogue} & \cellcolor{gray!15}{liberal} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{0.79}\\
jew & familial & christianStereotype & 0.98\\
\cellcolor{gray!15}{quran} & \cellcolor{gray!15}{dirty} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{1.12}\\
muslim & uneducated & muslimStereotype & 0.52\\
\cellcolor{gray!15}{torah} & \cellcolor{gray!15}{terrorist} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{0.93}\\
quran & hairy & jewStereotype & 1.18\\
\cellcolor{gray!15}{synagogue} & \cellcolor{gray!15}{violent} & \cellcolor{gray!15}{muslimStereotype} & \cellcolor{gray!15}{0.95}\\
bible & cheap & jewStereotype & 1.22\\
\cellcolor{gray!15}{christianity} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{0.97}\\
muslim & hairy & jewStereotype & 0.88\\
\cellcolor{gray!15}{islam} & \cellcolor{gray!15}{critical} & \cellcolor{gray!15}{christianStereotype} & \cellcolor{gray!15}{0.79}\\
muslim & conservative & christianStereotype & 0.45\\
\cellcolor{gray!15}{mosque} & \cellcolor{gray!15}{greedy} & \cellcolor{gray!15}{jewStereotype} & \cellcolor{gray!15}{1.15}\\
\bottomrule
\end{tabular}

@@

# Challenges to cosine-based bias metrics
\label{sec:challenges}

## Interpretability issues
\label{subsec:interpretability}

@@

In Figures  (\ref{fig:empirical0}-\ref{fig:empirical3}) we look at the empirical distributions, while paying attention to such divisions. The horizontal lines represent the values of 1-\textsf{MAC} (that is, we now talk in terms of cosine similarity rather than cosine distance; this is a trivial linear transformation) that the authors considered indicative of bias for stereotypes corresponding to  given  word lists. For instance, in religion, \textsf{MAC} was .859 and considered a sign of bias, so we plot $0\pm (1-.859)\approx .14$ lines around similarity = 0 (that is, distance = 1). Notice that most distributions are quite wide, and the proportions of even neutral or human neutral words with similarities higher than the value of 1-\textsf{MAC} deserving debiasing according to the authors is quite high.

```{r cosineDistributions6d,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "75%", warning = FALSE, message = FALSE}

@@

violinReligionReddit <- ggplot(religionReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+
  geom_vline(xintercept = .14) + geom_hline(yintercept = -.14, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 13)+ geom_hline(yintercept = .14, color = "grey")+
  labs(title =  "Religion (Reddit)")+
  geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "56%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "55%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "43%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "36%")+
  theme(plot.title.position = "plot",legend.position="top")

violinGenderReddit <- ggplot(genderReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  geom_vline(xintercept = .38) + geom_hline(yintercept = -.38, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 13)+ geom_hline(yintercept = .38, color = "grey")+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "96%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "87%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "53%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "47%")+
  labs(title =  "Gender (Reddit)")+
  theme(plot.title.position = "plot",legend.position='none')

violinRaceReddit <- ggplot(raceReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
  geom_violin(alpha = .5)+coord_flip()+geom_jitter(size = 0.5, alpha = .2, width=0.08)+
  geom_vline(xintercept = .11) + geom_hline(yintercept = -.11, color = "grey") +
  ylab("similarity")+theme_tufte(base_size = 13)+ geom_hline(yintercept = .11, color = "grey")+
   labs(title =  "Race (Reddit)")+
  stat_summary(fun.data = "mean_cl_boot", fun.args = list(
    conf.int = .95), geom = "pointrange",
               colour = "black", size  = .2  )+
  annotate(geom = "label", x = 4.2, y = 0, label = "46%")+
  annotate(geom = "label", x = 3.2, y = 0, label = "67%")+
  annotate(geom = "label", x = 2.2, y = 0, label = "39%")+
  annotate(geom = "label", x = 1.2, y = 0, label = "33%")+
  theme(plot.title.position = "plot",legend.position='none')

@@

\begin{figure}[H]
```{r fig:cosineDistributions8d,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning = FALSE, message = FALSE}

violinReligionReddit

\caption{Empirical distributions of cosine similarities  for the Religion word list  used in  the original paper. }

\label{fig:empirical0}
\end{figure}

\begin{figure}[H]
```{r fig:cosineDistributions8df,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%",  warning = FALSE, message = FALSE}

violinGenderReddit
```

\caption{Empirical distributions of cosine similarities for the Gender word list used in  the original paper.  }

\label{fig:empirical2}
\end{figure}

\begin{figure}[H]
```{r fig:cosineDistributions8dg,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning = FALSE, message = FALSE}

violinRaceReddit
```

\caption{Empirical distributions of cosine similarities  for the Race word list used in  the original paper.  }

\label{fig:empirical3}
\end{figure}

@@

##  Problems with pre-averaging
\label{subsec:problems}

The approaches we have been describing use  means of mean average cosine similarities to measure  similarity between protected words and attributes coming from harmful stereotypes. However, a closer inspection look at the individual values that are taken for the calculations, it turns out that there are quite a few  outliers and surprisingly dissimilar words. This problem becomes transparent when we examine the visualizations of the individual cosine distances, following the idea that one of the first steps in understanding data is to look at it. Let's start with inspecting two examples in Figures \ref{fig:muslim} and \ref{fig:priest}.

@@

visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 11)
```
\caption{Actual distances for the protected word "muslim". The most similar attributes tend to be the ones associated with it stereotypically, but then words associated with other stereotypes come closer than neutral or human predicates. }
\label{fig:muslim}
\end{figure}

\begin{figure}[H]
```{r visPriestd,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
religion <- read.csv("datasets/cleanedDatasets/religionRedditCleaned.csv")
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

@@

\caption{Actual distances for the protected word "priest", the situation is even less as expected: the nearest attributes are human attributes, and all there seems to be no clear pattern to the distances to other attributes.}
\label{fig:priest}
\end{figure}

@@

ourESplot <- ggplot(dataPLOT )+geom_histogram(aes(x=biasesD0, y = ..density..), bins = 50, alpha = .6)+
  labs(title="Distribution of effect sizes given a null model")+ xlab("effect size") +
  geom_vline(xintercept = ourEffectSize, linewidth = .5,
             lty = 2, alpha = .5)+theme_tufte(base_size = 13)+
  annotate("label", x = ourEffectSize, y = .06,
           label =  paste(round(ourEffectSize,2),
                          ", p=", 0.38, sep = ""), size = 4)+
    theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())

ourStatPlot <- ggplot(dataPLOT )+ geom_histogram(aes(x=svaluesD0, y = ..density..), alpha = .6, bins = 50) +
  labs(title="Distribution of test statistics given a null model")+ xlab("test statistic") +
  geom_vline(xintercept = ourStatistic, linewidth = .5,
             lty = 2, alpha = .5)+theme_tufte(base_size = 13)+
  annotate("label", x = ourStatistic, y = 1.2,
           label =  paste(round(ourStatistic,2),
                          ", p=", 0.01, sep = ""), size = 4)+
    theme(plot.title.position = "plot", axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())

@@

\begin{figure}[H]
```{r fig:ourDistancesPlot2e,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%", warning = FALSE, message = FALSE}

@@

\caption{Distributions of test statistics and effect sizes based on 10k simulations on the assumption of a null model in which all distances come from normal distribution with $\mu =0, \sigma = .08, n=10k$.}
\label{fig:ourDistances}
\end{figure}

@@

# Bayesian method
\label{sec:bayesian}

## Model construction
\label{subsec:model}

@@

## Posterior predictive check
\label{subsec:posterior}

@@

\begin{figure}[H]
```{r figposteriorPrCheckA6b,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%", fig.height= 6, dpi = 300, warning = FALSE, message = FALSE}
grid.arrange(posteriorPlotsA[[3]]+theme_tufte(base_size = 15),
             posteriorPlotsB[[3]] +theme_tufte(base_size = 15), ncol = 1)
```
\caption{Example of a posterior predictive check. (Left) Actual cosine distances are plotted against mean predictions with 89\% highest posterior density intervals. Notice that 90\% of actual values fall within the 89\% HPDI and 55\% of actual values fall into 50\% HPDI, which indicates appropriate performance of the model. The left-right alignment of differrent colors coressponds to the fact that cosine differences between elements of different categories differ, to some extent systematically (this will be studied in the results section). (Right) Densities of predicted and observed distances.}
\label{fig:posteriorCheck1}
\end{figure}

@@

# Results and discussion
\label{sec:results}

\vspace{1mm}
